Application of Machine Learning to Increase the Efficiency of the Global Search Algorithm for Solving Multicriterial Problems\thanks{This research was supported by the Russian Science Foundation, project No. 21-11-00204.}

### 2023-06-10 2023/NUMTA/Full paper/NUMTA_2023_MCO.tex
chqck the last part of the paper

In general case, the functions satisfying the Lipschitz condition are multiextremal, therefore, the convolution (\ref{eq:05}) will be multiextremal and Lipschitzian as well. It means that for solving the problem (\ref{eq:08}) it is required to apply methods of non-linear programming which provides finding the global optimum. There are many global optimization algorithms designed under different assumptions about properties of objective function; relevant references can be found, for instance, in monographs \cite{Pardalos2017,Marler2009,Sergeyev2013}. Some of them are theoretically substantiated and guarantee the convergence to global solution; the others are empirical, for example, nature-inspired methods that do not ensure finding the global optimum.

The information-statistical algorithms \cite{Gergel2017_1,Gergel2018,Barkalov2014,Barkalov2015,Barkalov2018,Strongin2003}  as deterministic methods with guaranteed convergence are among the most effective for analyzing multiextremal optimization problems. For solving multidimensional problems, ISA algorithms apply two approaches based on ideas of dimensionality reduction. In the framework of one approach the multidimensional problem is reduced to a family of univariate problems connected recursively (scheme of nested optimization) \cite{Grishagin2016,Zilinskas2015,Gergel2019_2}. The other approach \cite{Gergel2018,Barkalov2014,Sergeyev2013} exploits the known fact that $n$-dimensional box (1) and the closed interval $[0,1]$ of the real axis are equipotent sets and there exist unambiguous and continuous mappings of this interval onto the hyperparallelepiped (1) called \textit{Peano-type} curves or \textit{evolvents}.

Existence of Peano mappings can be applied to solving the optimization problems in the following way. Let there be a $n$-dimensional continuous function $g(y)$ defined in the domain (\ref{eq:01}) and $y(x)$ is a Peano-type curve. Then the continuity of $g(y)$ and $y(x)$ allows one to assert that
\begin{equation}
    \min_{y\in H} g(y) = \min_{x\in[0,1]} g(y(x))
\end{equation}
i.e., instead of minimization of the multidimensional function $g(y)$ in the domain (\ref{eq:01}) one can solve the minimization problem for the one-dimensional function $g(y(x))$ over interval $[0,1]$.

However, because of fundamental characteristics of the mapping $y(x)$ the reduced function $g(y(x))$ does not retain Lipschitz property in the case of Lipschitzian partial criteria from (\ref{eq:02}). In this situation the function $g(y(x))$ satisfies the H{\" o}lder condition 
\begin{equation}
    \label{eq:10}
    |g(y(x')) - g(y(x''))| \leq G \|x'-x''\|, x', x'' \in [0,1]
\end{equation}
where $G>0$ is H{\" o}lder constant. If the function $g(y)$ satisfies the Lipschitz property with a constant $L$ then the function $g(y(x))$ will satisfy the inequality (\ref{eq:10}) with $G=2L\sqrt{n+3}$ \cite{Strongin2003,Sergeyev2013}.

@@

Let the set $M$ contain the finite number $m$ of different weight vectors (\ref{eq:06a}):
\begin{equation}
    \label{eq:11}
    M=(\mu^1, \dots, \mu^m)
\end{equation}
where
\begin{equation}
    \label{eq:12}
    \mu^i=(\mu^i_k, \dots, \mu^i_k), 1 \leq i \leq m.
\end{equation}

The set $M$ generates a family   of scalar problems 
\begin{equation}
    \label{eq:13}
    W(\mu^i, y)= V(\mu^i, y) + \alpha \sum^k_{s=1}\mu^i_s \varphi_s (y) \to \min, y \in H, 1 \leq i \leq m.
\end{equation}

Let's denote as $f^i(x)$ the univariate function obtained in accordance with Peano mapping $y(x)$ from the function $W(\mu^i, y)$, i.e.,
\begin{equation}
    \label{eq:14}
    f^i(x) = W(\mu^i, y(x)), 1 \leq i \leq m.
\end{equation}

Minimization of a separate function $f^i(x)$ over interval $[0,1]$ allows one to find partial solution(s) of the initial MMCO problem (\ref{eq:03}) and solving all the problems
\begin{equation}
    \label{eq:15}
    f^i(x) \to min, x \in [0,1], 1 \leq i \leq m,
\end{equation}
enables to construct an approximation of the Pareto set.

The following computational scheme is proposed for solving the problems of the family (\ref{eq:15}). In this scheme, all the problems (\ref{eq:15}) are solved sequentially and during minimization of current function $f^i(x), i > 1$, the results of preceding problems $f^j(x) \to \min, 1 \leq j \leq i - 1$ completed already are taken into account.

@@

Hereinafter the term ``trial'' will denote the evaluation of the objective function $f^i(x)$ at a point $x^j \in [0,1]$ and obtaining the vector (result of trial)
\begin{equation}
    \label{eq:15a}
    \Omega^i(x^j)=(z^{i,j}, \omega^j_1, \dots, \omega^j_k),
\end{equation}
where $z^{i, j}=f^i(x^j)$, and the quantities
\begin{equation}
    \label{eq:16}
    \omega^i_s=\varphi_s(y(x^j)), 1 \leq s \leq k,
\end{equation}
are values of corresponding partial criteria at the point $x^j$. If a vector $M$ of weight coefficients (\ref{eq:11}) corresponds to the function (\ref{eq:14}) then 
\begin{equation}
    \label{eq:17}
    z^{i, j} = \max_{1\leq s \leq k} \mu_s \omega^j_s + \alpha \sum^k_{s=1} \omega^j_s.
\end{equation}

@@

Before starting the described algorithm to build a Pareto set approximation, the set (\ref{eq:11}) of weight coefficient vectors is supposed to be given. In the framework of the statement (\ref{eq:01})-(\ref{eq:03}) this set generates the family of $m$ multidimensional scalar problems (\ref{eq:13}) and corresponding to those the family of $m$ univariate problems (\ref{eq:13}).

@@

At the first stage when $f^i(x)=f^1(x)$, $q=2$ and $x^1=0, x^2=1$. These 2 trials generate results $\Omega^1(x^1), \Omega^1(x^2)$ from (\ref{eq:15a}) and initial information matrix $A^1_2$ from (\ref{eq:18}).

@@

\textbf{Rule 1.} Points $x^j, 1 \leq j\leq p,$ are ordered in increasing order and renumbered by subscripts, i.e.,
\begin{equation}
    \label{eq:19}
    x_0 = 0 < x_1 < \dots < x_{p-2} < x_{p-1} = 1.
\end{equation}

The values $u_j = f^i(x_j), 1\leq j \leq p$ are juxtaposed to the points $x_j, 1\leq j \leq p$.

\textbf{Rule 2.} Current estimation $\gamma$ of the H{\" o}lder constant  
\begin{equation}
    \label{eq:20}
\gamma=
\begin{cases}
  \begin{matrix}
     \gamma \Gamma & \Gamma >0 \\
     1 & \Gamma = 0 
  \end{matrix}
\end{cases}
\end{equation}
is computed, where
\begin{equation}
    \label{eq:21}
\Gamma=
\max_{1 \leq j \leq p-1} {\frac{|u_j - u_{j-1}|}{\rho_j} },
\end{equation}
\begin{equation}
    \label{eq:22}
\rho_j = \sqrt[n]{x_j-x_{j-1}},
\end{equation}
and $r>1$ is a parameter of the method.

@@

\textbf{Rule 7.} Neighboring points from (\ref{eq:19}) provide a partitioning of the interval $[0, 1]$ into $p-1$ subintervals $(x_{j-1}, x_j), 1 \leq j \leq p-1$.

A real value $B(j)$ is assigned to each subinterval $(x_{j-1}, x_j), 1 \leq j \leq p-1$. $B(j)$ is called \textit{the characteristic}  of this subinterval and is calculated as
\begin{equation}
    \label{eq:24}
    B(j) = \rho_j + \frac{(u_j-u_{j-1})^2}{\gamma^2 \rho_j} - \frac{2 (u_j+u_{j-1})}{\gamma} + \beta (d'_j + d'_{j-1})
\end{equation}
where $\beta > 0$ is a parameter influencing the contribution of the separating plane.

@@

\textbf{Rule 10.} If the inequality
\begin{equation}
    \label{eq:27}
    \rho_l=\sqrt[n]{x_l-x_{l-1}} < \varepsilon
\end{equation}
is true where $\varepsilon > 0$ is predefined accuracy, then the $i$-th  stage (minimization of function $f^i(x)$) is over.

@@

Computations were conducted on the supercomputer ``Lobachevsky'' of the Nizhni Novgorod State University (operating system CentOS 7, control system SLURM). One node of the supercomputer has 2 processors Intel Sandy Bridge E5-2660 2.2 GHz, 64 Gb RAM. One CPU contains 8 cores, i.e., 16 cores are available on one node. To obtain the executable program code the compilers Intel C++ 17.0.0 and python 3.9 were used. Computations were performed with the support of the package scikit-learn 0.24.2 and the applied optimization system Globalizer \cite{globalizerSystem}.

The experiment was organized in the following way. For efficiency evaluation of the proposed algorithm hereinafter called AGS-SP (Algorithm of Global Search with Separating Planes), it was compared with its modification AGS-R (Algorithm of Global Search – Reduced) where computational scheme is similar, but Rules 3- 6 are removed and in the characteristic (\ref{eq:24}) parameter $\beta = 0$, AGS-R ignores the information about the Pareto set approximation. Moreover, to expand the basis of the comparison, some known meta-heuristic evolutionary MMCO methods were involved into the experiment, namely, algorithms SMPSO \cite{NDG09}, OMOPSO \cite{RC05}, NSGA-II \cite{DPA02}, SPEA2 \cite{ZLT01}, IBEA \cite{ZK04} from the library jMetal \cite{DNA10}.

@@

Three series of experiments were executed. The first one dealt with a simple problem and was aimed at demonstration of positive influence of criteria space analysis by means of separating planes and at comparing with evolutionary techniques.

This problem was bi-criteria one \cite{CHIANDUSSI2012912}:
\begin{equation}
    \label{eq:28}
    \begin{matrix}
f_1 (y)=y_1 \\
f_2 (y)=(1+10y_2 )[1- \left(\frac{y_1}{1+10y_2} \right)^2- \frac{y_1}{1+10y_2} \sin(8 \pi y_1 ) ] \\
0 \leq y_1,y_2 \leq 1
    \end{matrix}
\end{equation}

@@

\begin{table}[ht]
\caption{Results of solving the MMCO problem (\ref{eq:28}).}
\label{tab:01}
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Algorithm} & \textbf{Number of trials} & \textbf{Number of Pareto points} & \textbf{HV}    \\ \hline
Pareto HV          &                           &                                  & \textbf{1.202} \\
AGS-R              & 300                       & 33                               & \textbf{1.146} \\
AGS-SP             & 226                       & 34                               & \textbf{1.146} \\
SMPSO              & 300                       & 15.75                            & 1.072          \\
OMOPSO             & 300                       & 15.89                            & 1.136          \\
NSGAII             & 300                       & 37.62                            & 0.913          \\
SPEA2              & 300                       & 38.32                            & 0.894          \\
IBEA               & 300                       & 41.99                            & 0.985          \\ \hline
\end{tabular}
\end{table}

The data of the table show that AGS-SP and AGS-R solve the problem (\ref{eq:28}) more efficient than evolutionary methods because while spending the same number of trials (and even less as AGS-SP) they provide the  better HV index and find more points in Pareto set approximation. If to compare AGS-SP and AGS-R with each other, then AGS-SP has an advantage due to HV indices being the same, but AGS-SP was "faster" as it executed fewer trials.

>>>>

In general case, the functions satisfying the Lipschitz condition are multiextremal, therefore, the convolution (\ref{eq:05}) will be multiextremal and Lipschitzian as well. It means that for solving the problem (\ref{eq:08}) it is required to apply methods of nonlinear programming which provides finding the global optimum. There are many global optimization algorithms designed under different assumptions about properties of objective function; relevant references can be found, for instance, in monographs \cite{Pardalos2017,Marler2009,Sergeyev2013}. Some of them are theoretically substantiated and guarantee the convergence to global solution; the others are empirical, for example, nature-inspired methods that do not ensure finding the global optimum.

The information-statistical algorithms \cite{Gergel2017_1,Gergel2018,Barkalov2014,Barkalov2015,Barkalov2018,Strongin2003}  as deterministic methods with guaranteed convergence are among the most effective for analyzing multiextremal optimization problems. For solving multidimensional problems, these algorithms apply two approaches based on ideas of dimensionality reduction. In the framework of one approach the multidimensional problem is reduced to a family of univariate problems connected recursively (scheme of nested optimization) \cite{Grishagin2016,Zilinskas2015,Gergel2019_2}. The other approach \cite{Gergel2018,Barkalov2014,Sergeyev2013} exploits the known fact that $n$-dimensional box (1) and the closed interval $[0,1]$ of the real axis are equipotent sets and there exist unambiguous and continuous mappings of this interval onto the hyperparallelepiped (1) called \textit{Peano-type curves} or \textit{evolvents}.

Existence of Peano mappings can be applied to solving the optimization problems in the following way. Let there be a $n$-dimensional continuous function $g(y)$ defined in the domain (\ref{eq:01}) and $y(x)$ is a Peano-type curve. Then the continuity of $g(y)$ and $y(x)$ allows one to assert that
\begin{equation}
    \min_{y\in H} g(y) = \min_{x\in[0,1]} g(y(x)),
\end{equation}
i.e. instead of minimization of the multidimensional function $g(y)$ in the domain (\ref{eq:01}) one can solve the minimization problem for the one-dimensional function $g(y(x))$ over the interval $[0,1]$.

However, because of fundamental characteristics of the mapping $y(x)$ the reduced function $g(y(x))$ does not retain Lipschitz property in the case of Lipschitzian partial criteria from (\ref{eq:02}). In this situation the function $g(y(x))$ satisfies the H{\" o}lder condition 
\begin{equation}
    \label{eq:10}
    |g(y(x')) - g(y(x''))| \leq G \left|x'-x''\right|^{1/n}, \; x', x'' \in [0,1],
\end{equation}
where $G>0$ is H{\" o}lder constant. If the function $g(y)$ satisfies the Lipschitz property with a constant $L$ then the function $g(y(x))$ will satisfy the inequality (\ref{eq:10}) with $G=2L\sqrt{n+3}$ \cite{Strongin2003,Sergeyev2013}.

@@

Let the set $M$ contain the finite number $m$ of different weight vectors (\ref{eq:06a}):
\begin{equation}
    \label{eq:11}
    M=(\mu^1, \dots, \mu^m),
\end{equation}
where
\begin{equation}
    \label{eq:12}
    \mu^i=(\mu^i_k, \dots, \mu^i_k), \; 1 \leq i \leq m.
\end{equation}

The set $M$ generates a family   of scalar problems 
\begin{equation}
    \label{eq:13}
    W(\mu^i, y)= V(\mu^i, y) + \alpha \sum^k_{s=1}\mu^i_s \varphi_s (y) \to \min, \; y \in H, 1 \leq i \leq m.
\end{equation}

Let's denote as $f^i(x)$ the univariate function obtained in accordance with Peano mapping $y(x)$ from the function $W(\mu^i, y)$, i.e.,
\begin{equation}
    \label{eq:14}
    f^i(x) = W(\mu^i, y(x)), \; 1 \leq i \leq m.
\end{equation}

Minimization of a separate function $f^i(x)$ over interval $[0,1]$ allows one to find partial solution(s) of the initial MMCO problem (\ref{eq:03}) and solving all the problems
\begin{equation}
    \label{eq:15}
    f^i(x) \to min, \; x \in [0,1], \; 1 \leq i \leq m,
\end{equation}
enables to construct an approximation of the Pareto set.

The following computational scheme is proposed for solving the problems of the family (\ref{eq:15}). In this scheme, all the problems (\ref{eq:15}) are solved sequentially and during minimization of current function $f^i(x), i > 1$, the results of preceding problems $f^j(x) \to \min, 1 \leq j \leq i - 1,$ completed already are taken into account.

@@

Hereinafter the term ``trial'' will denote the evaluation of the objective function $f^i(x)$ at a point $x^j \in [0,1]$ and obtaining the vector (trial result)
\begin{equation}
    \label{eq:15a}
    \Omega^i(x^j)=(z^{i,j}, \omega^j_1, \dots, \omega^j_k),
\end{equation}
where $z^{i, j}=f^i(x^j)$, and the quantities
\begin{equation}
    \label{eq:16}
    \omega^i_s=\varphi_s(y(x^j)), 1 \leq s \leq k,
\end{equation}
are values of corresponding partial criteria at the point $x^j$. If a vector $M$ of weight coefficients (\ref{eq:11}) corresponds to the function (\ref{eq:14}) then 
\begin{equation}
    \label{eq:17}
    z^{i, j} = \max_{1\leq s \leq k} \mu_s \omega^j_s + \alpha \sum^k_{s=1} \omega^j_s.
\end{equation}

@@

Before starting the described algorithm to build a Pareto set approximation, the set (\ref{eq:11}) of weight coefficient vectors is supposed to be given. In the framework of the statement (\ref{eq:01})--(\ref{eq:03}) this set generates the family of $m$ multidimensional scalar problems (\ref{eq:13}) and corresponding to those the family of $m$ univariate problems (\ref{eq:15}).

@@

At the first stage when $f^i(x)=f^1(x)$, $q=2$ and $x^1=0, x^2=1$. These two trials generate results $\Omega^1(x^1), \Omega^1(x^2)$ from (\ref{eq:15a}) and initial information matrix $A^1_2$ from (\ref{eq:18}).

@@

\textbf{Rule 1.} Points $x^j, 1 \leq j\leq p,$ are ordered in increasing order and renumbered by subscripts, i.e.,
\begin{equation}
    \label{eq:19}
    x_0 = 0 < x_1 < \dots < x_{p-2} < x_{p-1} = 1.
\end{equation}
The values $u_j = f^i(x_j), 1\leq j \leq p,$ are juxtaposed to the points $x_j, 1\leq j \leq p$.

\textbf{Rule 2.} Current estimation $\gamma$ of the H{\" o}lder constant  
\begin{equation}
    \label{eq:20}
\gamma=
\begin{cases}
  \begin{matrix}
     \gamma \Gamma, & \Gamma >0 \\
     1, & \Gamma = 0 
  \end{matrix}
\end{cases}
\end{equation}
is computed, where
\begin{equation}
    \label{eq:21}
\Gamma=
\max_{1 \leq j \leq p-1} {\frac{|u_j - u_{j-1}|}{\rho_j} },
\end{equation}
\begin{equation}
    \label{eq:22}
\rho_j = \sqrt[n]{x_j-x_{j-1}},
\end{equation}
and $r>1$ is a parameter of the method.

@@

\textbf{Rule 7.} Neighboring points from (\ref{eq:19}) provide a partitioning of the interval $[0, 1]$ into $p-1$ subintervals $(x_{j-1}, x_j), 1 \leq j \leq p-1$.
A real value $B(j)$ is assigned to each subinterval $(x_{j-1}, x_j), 1 \leq j \leq p-1$. $B(j)$ is called \textit{the characteristic}  of this subinterval and is calculated as
\begin{equation}
    \label{eq:24}
    B(j) = \rho_j + \frac{(u_j-u_{j-1})^2}{\gamma^2 \rho_j} - \frac{2 (u_j+u_{j-1})}{\gamma} + \beta (d'_j + d'_{j-1}),
\end{equation}
where $\beta > 0$ is a parameter influencing the contribution of the separating plane.

@@

\textbf{Rule 10.} If the inequality
\begin{equation}
    \label{eq:27}
    \rho_l=\sqrt[n]{x_l-x_{l-1}} < \varepsilon
\end{equation}
is true, where $\varepsilon > 0$ is predefined accuracy, then the $i$-th  stage (minimization of function $f^i(x)$) is over.

@@

Computations were conducted on the supercomputer ``Lobachevsky'' of the Nizhni Novgorod State University (operating system CentOS 7, control system SLURM). One node of the supercomputer has 2 processors Intel Sandy Bridge E5-2660 2.2 GHz, 64 Gb RAM. One CPU contains 8 cores, i.e., 16 cores are available on one node. To obtain the executable program code the compilers Intel C++ 17.0.0 and python 3.9 were used. Computations were performed with the support of the package scikit-learn 0.24.2 and the optimization software Globalizer \cite{globalizerSystem}.

The experiment was organized in the following way. For efficiency evaluation of the proposed algorithm hereinafter called AGS-SP (Algorithm of Global Search with Separating Planes), it was compared with its modification AGS-R (Algorithm of Global Search Reduced) where computational scheme is similar, but Rules 3 -- 6 are removed and in the characteristic (\ref{eq:24}) parameter $\beta = 0$. AGS-R ignores the information about the Pareto set approximation. Moreover, to expand the basis of the comparison, some known meta-heuristic evolutionary MMCO methods were involved into the experiment, namely, algorithms SMPSO \cite{NDG09}, OMOPSO \cite{RC05}, NSGA-II \cite{DPA02}, SPEA2 \cite{ZLT01}, IBEA \cite{ZK04} from the jMetal framework \cite{DNA10}.

@@

Three series of experiments were executed. The first one dealt with a simple problem and was aimed at demonstration of positive influence of criteria space analysis by means of separating planes and at comparing with evolutionary techniques.
This problem was bi-criteria one \cite{CHIANDUSSI2012912}:
\begin{equation}
    \label{eq:28}
    \begin{matrix}
f_1 (y)=y_1, \\
f_2 (y)=(1+10y_2 )[1- \left(\frac{y_1}{1+10y_2} \right)^2- \frac{y_1}{1+10y_2} \sin(8 \pi y_1 ) ], \\
0 \leq y_1,y_2 \leq 1.
    \end{matrix}
\end{equation}

@@

\begin{table}[ht]
\caption{Results of solving the MMCO problem (\ref{eq:28}).}
\label{tab:01}
\center
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Algorithm} & \textbf{Number of trials} & \textbf{Number of Pareto points} & \textbf{HV}    \\ \hline
Pareto HV          & ---                       & ---                              & \textbf{1.202} \\
AGS-R              & 300                       & 33                               & \textbf{1.146} \\
AGS-SP             & 226                       & 34                               & \textbf{1.146} \\
SMPSO              & 300                       & 15.75                            & 1.072          \\
OMOPSO             & 300                       & 15.89                            & 1.136          \\
NSGAII             & 300                       & 37.62                            & 0.913          \\
SPEA2              & 300                       & 38.32                            & 0.894          \\
IBEA               & 300                       & 41.99                            & 0.985          \\ \hline
\end{tabular}
\end{table}

The data of the table show that AGS-SP and AGS-R solve the problem (\ref{eq:28}) more efficient than evolutionary methods because while spending the same number of trials (and even less as AGS-SP) they provide the  better HV index and find more points in Pareto set approximation. If to compare AGS-SP and AGS-R with each other, then AGS-SP has an advantage due to HV indices being the same, but AGS-SP was ``faster'' as it executed fewer trials.
